\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2017-04-24 doc/publications/plos16/nm.tex
rewrite

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
\setlength{\headheight}{27.023pt}
\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\sf PLOS}

@@

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{\cursedforest - A Random Forest Implementation for ``Big'' and ``Wide'' Data} % Please use "title case" (capitalize all terms in the title except conjunctions, prepositions, and articles).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Aidan O'Brien\textsuperscript{1,2},
Piotr Szul\textsuperscript{3},
Robert Dunne\textsuperscript{4}, 
Paul Leo \textsuperscript{5},
Emma L. Duncan \textsuperscript{5,6,7},
Natalie Twine\textsuperscript{1,8}
Stephanie Li\textsuperscript{9},
James Doecke\textsuperscript{9},
Nick Ellis\textsuperscript{10}, and
Denis C. Bauer\textsuperscript{1,*}
%, with the Lorem Ipsum Consortium\textsuperscript{\textpilcrow}
\\
\bigskip
\textbf{1} Health \& Biosecurity, CSIRO, Sydney, NSW, Australia
\\
\textbf{2} ANU, Canberra, Australia
\\
\textbf{3} Data61, CSIRO, Brisbane, QLD, Australia
\\
\textbf{4} Data61, CSIRO, Sydney, NSW, Australia
\\
\textbf{5} Institute of Health and Biomedical Innovation, Queensland University of Technology, Brisbane, QLD, Australia
\\
\textbf{6} Department of Endocrinology, Royal Brisbane and Women's Hospital, Brisbane, QLD, Australia
\\
\textbf{7} Faculty of Medicine and Biomedical Sciences, University of Queensland, Brisbane, QLD, Australia
\\
\textbf{8} MQ, QLD, Australia
\\
\textbf{9} Health \& Biosecurity, CSIRO, Brisbane, QLD, Australia
\\
\textbf{10} Oceans \& Atmosphere, CSIRO, Brisbane, QLD, Australia
\bigskip

@@

Analyses are performed on ever larger datasets with unprecedentedly detailed information per data point. Specifically in
genomics, cohort sizes are in the thousands, each comprising of millions of variants that make up a person's genetic
differences. Traditional approaches are unable to cater for this task and even standard implementations using novel
compute paradigms, such as Hadoop/Spark, are limited by the finite resource of computer memory on such
data. Specifically, Spark ML's supervised machine learning methods suffer from "the curse of dimensionality". That is,
they scale well with samples ($n$) but not with features ($p$, variants).

@@

Identifying the relevant or disease specific base pair differences between individuals is the focus of genome wide
association studies (GWAS).  These analyses are typically performed on only the most frequently differing base pairs,
termed single nucleotide polymorphisms (SNPs), by applying linear or logistic regression analysis to each SNP
separately~\cite{CCC2007}.  It has since been demonstrated that not taking interactions between SNPs into account
when filtering for potential disease mutations is inadequate~\cite{Manolio2009,Yang2011}.

@@

Despite overcoming the node-boundary limitation, the standard implementation of random forest in Spark ML
is not able to handle the extremely ``wide'' genomic data as it was developed for a large number of samples
with only modest dimensionality~\cite{NIPS2016_6366}.  Although Spark ML can build a random forest model on a subset of the data (chromosome
1), we show that the time taken is excessive due to the large amount of data being aggregated and processed by the driver node during
intermediate stages of building the model.  This unbalanced work load where the driver
node becomes the bottleneck and worker nodes are idle prevents a seamless scaling to larger datasets. We also show that the memory
requirements per executor increases with dimensionality due to the data types Spark ML uses.%, which we elaborate on in the discussion.

@@

\cite{Biau.2012} provides a proof that a random forest model will converge at a rate that depends on
the cardinality of the set $S$ of ``strong predictors'' rather than on the number of variables $p$. That is, given a
function $y=f(S)$ depending on a set of variables $S$, a random forest will converge to the true function $y$ with a
rate that depends on the size of $S$ and not the number of noise variables in the dataset. If the size of $S$ is small
compared to the total number of variables, then the rate of convergence will be quicker. It is not clear what the size
of the set $S$ will be in a GWAS application. There are cases of small groups of SNPs with large effects and also of
conditions caused by the interactions of larger groups of SNPs.

@@

However, within an RDD, Spark ML stores each sample as a vector. Unlike RDDs, which can be partitioned and distributed
across multiple nodes, each vector must be present in its entirety on any node accessing it. This is no problem with
typical datasets; however, as dimensionality increases, the vectors eventually reach a size where they can no longer fit
into a single node's memory.

@@

\begin{table}[!ht]
%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\begin{flushleft} 
\centering
\caption{\bf Possible encodings of GWAS data \cite{Goldstein.et.al.2011}}
\begin{tabular}{|l|l|l|l||}
\hline
{\bf Type} & {\bf Mechanism}                                         & {\bf Partition} \\ 
\thickhline
Additive   & Each additional minor allele increases variation        & 0, 1, 2         \\
Dominant   & Presence of at least 1 minor allele increases variation & 0, $\{1,2\}$          \\
Recessive  & Two minor alleles needed for variation                  & $\{0,1\}$, 2          \\
Heterosis  & Heterozygote leads to variation                         & $\{0,2\}$, 1          \\ \hline
%\end{adjustwidth}
\end{tabular}
\label{table:encodings}
\end{flushleft}
\end{table}

@@

\subsection{Parameter settings}
We consider the parameter settings for the random forest algorithm. We use the R notation from the \randomforest
package \cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate
the advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience:
\begin{itemize}
\item \ntree{}  -- the number of trees.  The number of trees necessary for good performance grows with the number of
  predictors.  \cite{Liaw.and.Weiner.2002} suggest that a high \ntree is necessary to get stable estimates of variable
  importance and proximity; however, even though the variable importance measures may vary from run to run, we note that
  it is possible for a random forest model to have a poorer fit and still have an accurate ranking of variable
  importance;
\item \mtry{}  -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry may give
  better performance;
\item the size and complexity of the individual trees is controlled in \randomforest by setting \texttt{nodesize}, the
  minimum size of terminal nodes. It is controlled in Spark ML by setting \texttt{maxDepth}, the maximum depth of each
  tree in the forest.\footnote{The Spark ML documentation \cite{Spark.2016} sets \texttt{maxDepth} to 4 in their
    classification example. This may give an estimate with a higher bias and, as such, may be a poor choice. See the
    discussion in \cite{Dietterich.2002} where it is shown that the question of reducing bias versus variance is more
    complicated than first thought. They show that bagging may reduce bias as well as variance but give a
    better result for low bias learners. \cite{Dietterich.2002} notes that if the bootstrap replicate approximation were
    correct (i.e.~if the bootstrap sample came from an identical distribution to the data), then bagging would reduce
    variance without changing bias.}
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 1:  Feature Selection and classification on synthetic data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest can evaluate variable importance for millions of features}
\label{section:synthetic}
In this section we demonstrate the two features of \cursedforest, namely: variable selection analysis to select the most
associated set of features amongst millions of dimensions; and its ability to accurately classify the labels of
high-dimensional samples.  We therefore generate a synthetic dataset with 2.5 million variables ($p=2,500,000$), of which 5 
are related to the response variable, and 5000 samples ($n=5000$), as discussed in the methods section.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 2: Feature Selection on real data 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{GWAS-style analysis using \cursedforest\ identifies novel variants associated with Bone Mineral Density}
In this section we use \cursedforest\ to perform feature selection on over 7.2 million genomic variants and identify the
locations associated with Bone Mineral Density (BMD)~\cite{Duncan.2011}. 
%We build a model on the BMD dataset as described in Methods using a binary high or low BMD value as the label.  
We rank variants by their Gini impurity, with the most important one ranked at position 1, and return the top 100,000 variants.

@@

As shown in Figure~\ref{figure:knowngeneranks}, the top 30 features are dominated by variants from five well established BMD genes; TNFRSF11B (TNF Receptor Superfamily Member 11b or Osteoprotegerin), CTNNB (Catenin Beta 1), ZBTB40 (Zinc Finger And BTB Domain-Containing Protein 40), IBSP (Integrin Binding Sialoprotein) and MEF2C (Myocyte Enhancer Factor 2C). 
The BMD association has been independently replicated for four of the five genes: CTNNB1~\cite{Pei2016, Rivadeneira2009}, MEF2C~\cite{Pei2016a}, TNFRSF11B~\cite{Rivadeneira2009,Zhang2014,Kemp2014,Paternoster2013} and ZBTB40~\cite{Rivadeneira2009,  Zhang2014,Nielson2016}. The association of IBSP has not been replicated in other GWAS studies, but homozygous knockouts of this gene have a known skeletal phenotype in mice. Mice deficient in IBSP have impaired bone growth and mineralization, with dramatically reduced bone formation~\cite{Malaval2008}. CTNBB1, MEF2C and TNFRSF11B also have known skeletal phenotypes. CTNNB1 is a key downstream member of the Wnt signaling pathway, and deletion of CTNNB1 in mice results in severe osteopenia, while constitutive activation of CTNNB1 results in massively increased bone deposition~\cite{Holmen2005}. MEF2C is a transcription factor which controls bone mass by regulating osteoclastic bone resorption, where mice with a MEF2C deletion in osteocytes had increased bone mass compared with normal controls~\cite{Kramer2012}. TNFRSF11B (or OPG) is a secreted glycoprotein which regulates bone mass~\cite{Simonet1997}. ZBTB40 has no known biological or molecular involvement in skeletal phenotypes~\cite{Styrkarsdottir2008}, even though it has been associated with BMD in multiple GWAS and is expressed in bone.
SNPs in linkage disequilibrium (LD) with each other at these loci also show a high importance score, as exemplified by TNFRSF11B in Figure 3b (Supplemental Figure 2 and 3 for the other genes). %~\ref{manhattenzoom}).

@@

In contrast, one of the lowest ranked gene with literature implied BMD association is DCDC5 (Doublecortin Domain Containing 5) ranked 23,842
(see Figure 2).
%~\ref{figure:knowngeneranks}).  
Similar to the conclusions drawn by Duncan {\it et
  al.}~\cite{Duncan.2011}, our results also indicate that this gene does not have a strong association with BMD. DCDC5 is not highly expressed in bone, and any mechanisms by which this gene might regulate BMD are unclear
\cite{Thakker2012}.  Although variants near DCDC5 are associated with BMD \cite{Rivadeneira2009}, it is questionable as
to whether the gene itself contributes to BMD.
Similarly, the lowest ranked gene with annotated BMD links, MAPT (microtubule associated protein tau) at 36,262, is primarily expressed in the nervous system and has not been identified in bone tissue. 
It is therefore likely that the association is through a co-morbidity with Alzheimer's disease~\cite{Dengler-Crish2017} and not a major contributing factor in this cohort.

@@

There are also four novel locations amongst the top 30 most associated variants as annotated in Figure 3a.
%~\ref{figure:rbo-prod.png}. 
Most notably a lincRNA (LINC01006) on chromosome 7, which has been associated with serum creatinine levels in two independent GWAS studies~\cite{Pattaro2010,Chambers2010}. 
Subjects with low serum creatinine levels were at a higher risk for low BMD~\cite{Huh2015} and in a rat study high creatine monohydrate diet resulted in greater lumbar BMD~\cite{Antolic2007}, 
with creatinine being the breakdown product of creatine. The variants overlapping the 5' end of the lincRNA are the highest ranked features in this region and all LD SNPs also show high importance (see Fig 3d). Also amonst the top 30 features is PRODH (proline dehydrogenase) located on 22q11, a loci observed by Stagi {\it
  et al.}~\cite{Stagi.2010} to significantly reduce bone mass in individuals with a micro-deletion in this region. 
PRODH catalyses the first step in proline degradation, and as proposed by Phang {\it et al.}~\cite{Phang.2008}, proline is made available by collagen degradation. 
Again, all LD SNPs also show high importance (see Figure 3c). %~\ref{zoomplot}). 
%Ranked at position 5 (rs6961378) and 18 (rs146034056) are located on chromosome 7 overlapping the lincRNA (LINC01006), which as been linked to vertebrate development~\cite{Woolfe2005}.
In contrast rs17195090 and rs112558362 are single feature loci (Supplementary Figure 4) hence likely due to erroneous imputation or genotyping.

@@

\cursedforest\ can successfully perform GWAS-type analyses replicating previous findings and uncovering novel
associations due to random forests also modelling interacting features.

@@

\begin{figure}[tbhp]
  \begin{adjustwidth}{-2.00in}{0in}
    \caption{\textbf{Rank of literature assigned BMD genes}}
    \label{figure:knowngeneranks}
    \includegraphics[totalheight=6.5cm]{./figs/BMDgenes_landscape.pdf}
    \begin{flushleft}
      The figure shows the rank of the 100 literature assigned BMD genes. Each ball represents a SNP assigned to the closest gene with colour representing the distance (max 100 kb). The size shows the Gini impurity score with larger balls representing more important SNPs. The horizontal line separates the genes with SNPs amongst the Top 20 associated features. 
     \end{flushleft}
  \end{adjustwidth}
\end{figure}

\begin{figure}[tbhp]
  \begin{adjustwidth}{-2.00in}{0in}
    \caption{\textbf{Genome-wide importance of SNPs and visualisation of the 22q.11 region}}
    \label{figure:ranksumtest}
     \begin{subfigure}[b]{0.5\linewidth}
     \includegraphics[totalheight=7cm]{./figs/BMDTopWholeGenome.png}
      \label{figure:rbo-prod.png} 
    \end{subfigure} 
 \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[totalheight=7cm]{./figs/new_10k_TNFRSF11B.pdf}
      \label{zoomplotTNFRSF} 
   \end{subfigure} 
    \begin{subfigure}[b]{0.5\linewidth}
     \includegraphics[totalheight=7cm]{./figs/new_10k_PRODH.pdf}
      \label{zoomplotprodh} 
    \end{subfigure} 
        \begin{subfigure}[b]{0.5\linewidth}
     \includegraphics[totalheight=7cm]{./figs/EUR_chr7_156_2-156_6mb.pdf}
      \label{zoomplotlinc} 
    \end{subfigure} 
    \begin{flushleft}
      (a) The figure shows genome-wide SNPs and their importance. Highlighted are the 100 literature assigned BMD genes (green) and four novel loci. The horizontal line indicates the top 30 features. 
      Locus Zoomplot for (b) the top ranked gene TNFRSF11B and the two putative novel associations (c) PRODH and (e) LINC01006. Each point represents a variant with the normalised Gini impurity score on the left y axis and LD to the central SNP in colour with the highest $r^2$ in red.
     \end{flushleft}
  \end{adjustwidth}
        \label{manhattenzoom} 
\end{figure}

@@

@@

We have demonstrated that using a different parallelization model can extend random forests to the case of an
extremely large number of variables. We have treated the case of variable selection in a $p >> n$ model, where most of
the variables are uninformative, and have demonstrated the utility of the model for large GWAS datasets. By comparing
this implementation to other implementations (including those optimized for large datasets) we have demonstrated the
utility of this approach.

Donoho and Tanner \cite{Donoho.and.Tanner.2009} give a ``universal phase change'' result that has applications in a
large number of areas, including variable selection in high dimensions. The theoretical phase change boundary is based on arguments
from combinatorial geometry.

They argue that there is a sharp disjunction between the cases where informative variables may be recovered with a high
accuracy by procedures like stepwise variable selection, and the cases where they can not be recovered. The boundary is
shown in Fig~\ref{figure:phase-diagram-equivalence.png} in a space parameterized by the level of underdeterminatio,n
$\delta = n/p$, and by the sparsity, $\rho =k/n$ (where $k$ is the number of informative variables).  Above the
phase-transition line variable recovery is still possible by a combinatorial approach such as all-subsets
variable selection.

\begin{figure}[tbhp] 
 \begin{adjustwidth}{-1.00in}{0in}
    \centering
    \includegraphics[totalheight=6cm]{./figs/phase2.png} 
    \caption{{\bf The phase change diagram.}}
    \label{figure:phase-diagram-equivalence.png} 
    \vspace{4ex}
  \end{adjustwidth}
\end{figure}

>>>>

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
\setlength{\headheight}{27.023pt}

%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
%\lfoot{\sf PLOS}

@@

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{\cursedforest - A Random Forest Implementation for ``Big'' and ``Wide'' Data} % Please use "title case" (capitalize all terms in the title except conjunctions, prepositions, and articles).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Aidan O'Brien\textsuperscript{1,2},
Piotr Szul\textsuperscript{3},
Robert Dunne\textsuperscript{4}, 
Paul Leo \textsuperscript{5},
Emma L. Duncan \textsuperscript{5,6,7},
Natalie A. Twine\textsuperscript{1,8}
Stephanie Li\textsuperscript{9},
James Doecke\textsuperscript{9},
Nick Ellis\textsuperscript{10}, and
Denis C. Bauer\textsuperscript{1,*}
%, with the Lorem Ipsum Consortium\textsuperscript{\textpilcrow}
\\
\bigskip
\textbf{1} Health \& Biosecurity, CSIRO, Sydney, NSW, Australia
\\
\textbf{2} ANU, Canberra, Australia
\\
\textbf{3} Data61, CSIRO, Brisbane, QLD, Australia
\\
\textbf{4} Data61, CSIRO, Sydney, NSW, Australia
\\
\textbf{5} Institute of Health and Biomedical Innovation, Queensland University of Technology, Brisbane, QLD, Australia
\\
\textbf{6} Department of Endocrinology, Royal Brisbane and Women's Hospital, Brisbane, QLD, Australia
\\
\textbf{7} Faculty of Medicine and Biomedical Sciences, University of Queensland, Brisbane, QLD, Australia
\\
\textbf{8} MQ, QLD, Australia
\\
\textbf{9} Health \& Biosecurity, CSIRO, Brisbane, QLD, Australia
\\
\textbf{10} Oceans \& Atmosphere, CSIRO, Brisbane, QLD, Australia
\bigskip

@@

Analyses are performed on ever larger datasets with unprecedentedly detailed information per data point. Specifically in genomics, cohort sizes are in the thousands, each comprising of Millions of variants that make up a person's genetic differences. Traditional approaches are unable to cater for this task and even standard implementations using novel compute paradigms, such as Hadoop/Spark, are limited by the finite resource of computer memory on such data. Specifically, Spark ML's supervised machine learning methods suffer from "the curse of dimensionality". That is, they scale well with samples (n) but not with features (p, variants).

@@

Identifying the relevant or disease specific base pair differences between individuals is the focus of genome wide
association studies (GWAS).  These analyses are typically performed on only the most frequently differing base pairs,
termed single nucleotide polymorphisms (SNPs), by applying linear or logistic regression analysis to each SNP
separately~\cite{CCC2007}.  It has since been demonstrated that not taking interactions between SNPs into account
when filtering for potential disease mutations is inadequate~\cite{Manolio2009,Yang2011} and the approach should 
in-fact be extended to interrogating all genomic positions[CITATION WES vs WGS?].

@@

Despite overcoming the node-boundary limitation, the standard implementation of random forest in Spark ML
is not able to handle the extremely ``wide'' genomic data as it was developed for a large number of samples
with only modest dimensionality~\cite{NIPS2016_6366}.  Although Spark ML can build a random forest model on a subset of the data (chromosome
1), we show that the time taken is excessive due to the large amount of data being aggregated and processed by the driver node during
intermediate stages of building the model.  This unbalanced work load where the driver
node becomes the bottleneck and worker nodes being idle prevents a seamless scaling to larger datasets. We also show that the memory
requirements per executor increases with dimensionality due to the data types Spark ML uses.%, which we elaborate on in the discussion.

@@

\cite{Biau.2012} provides a proof that a random forest model will converge at a rate that depends on
the cardinality of the set $S$ of ``strong predictors'' rather than on the number of variables $p$. That is, given a
function $y=f(S)$ depending on a set of variables $S$, a random forest will converge to the true function $y$ with a
rate that depends on the size of $S$ and not the number of noise variables in the dataset. If the size of $S$ is small
compared to the total number of variables, then the rate of convergence will be quicker. It is not clear what the size
of the set $S$ will be in a GWAS application. There are cases of small groups of SNPs with large effects and also of
conditions caused by the interactions of large groups of SNPs.

@@

However, within an RDD, Spark ML stores each sample as a vector. Unlike RDDs, which can be partitioned and distributed
across multiple nodes, each vector must be present in its entirety on any node accessing it. This is no problem with
typical datasets, however as dimensionality increases, the vectors eventually reach a size where they can no longer fit
into a single node's memory.

@@

\begin{table}[!ht]
%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\begin{flushleft} 
\centering
\caption{\bf Possible encodings of GWAS data \cite{Goldstein.et.al.2011}}
\begin{tabular}{|l|l|l|l||}
\hline
{\bf Type} & {\bf Mechanism}                                         & {\bf Partition} \\ 
\thickhline
Additive   & Each additional minor allele increases variation        & 0, 1, 2         \\
Dominant   & Presence of at least 1 minor allele increases variation & 0, 1/2          \\
Recessive  & Two minor alleles needed for variation                  & 0/1, 2          \\
Heterosis  & Heterozygote leads to variation                         & 0/2, 1          \\ \hline
%\end{adjustwidth}
\end{tabular}
\label{table:encodings}
\end{flushleft}
\end{table}

@@

\subsection{Parameter settings}
We consider the parameter settings for the random forest algorithm. We use the R notation from the \randomforest
package \cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate
the advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience:
\begin{itemize}
\item \ntree{}  -- the number of trees.  The number of trees necessary for good performance grows with the number of
  predictors.  \cite{Liaw.and.Weiner.2002} suggest that a high \ntree is necessary to get stable estimates of variable
  importance and proximity; however, even though the variable importance measures may vary from run to run, we note that
  it is possible for a random forest model to have a poorer fit and still have an accurate ranking of variable
  importance;
\item \mtry{}  -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry may give
  better performance;
\item the size and complexity of the individual trees is controlled in \randomforest by setting \texttt{nodesize}, the
  minimum size of terminal nodes. It is controlled in Spark ML by setting \texttt{maxDepth}, the maximum depth of each
  tree in the forest.\footnote{The Spark ML documentation \cite{Spark.2016} sets \texttt{maxDepth} to 4 in their
    classification example. This may give an estimate with a higher bias and, as such, may be a poor choice. See the
    discussion in \cite{Dietterich.2002} where it is shown that bagging may reduce bias as well as variance but give a
    better result for low bias learners. \cite{Dietterich.2002} notes that if the bootstrap replicate approximation were
    correct (i.e.~if the bootstrap sample came from an identical distribution to the data), then bagging would reduce
    variance without changing bias.}
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 2: Feature Selection on real data 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{GWAS-style analysis using \cursedforest\ identifies novel variants associated with Bone Mineral Density}
\subsection{GWAS-style analyses using \cursedforest\ faithfully identify known disease-accociated genes}
In this section we use \cursedforest\ to perform feature selection on over 7.2 million genomic variants and identify the
genomic locations associated with Bone Mineral Density (BMD)~\cite{Duncan.2011}. 
%We build a model on the BMD dataset as described in Methods using a binary high or low BMD value as the label.  
We rank variants by their Gini impurity, with the most important one ranked at position 1, and return the top 100,000 variants.

@@

%~\ref{figure:BMDTopWholeGenome}, 
As shown in Figure 1a, the top 30 features are dominated by variants from five well established BMD genes; TNFRSF11B (TNF Receptor Superfamily Member 11b or Osteoprotegerin), CTNNB (Catenin Beta 1), ZBTB40 (Zinc Finger And BTB Domain-Containing Protein 40), IBSP (Integrin Binding Sialoprotein) and MEF2C (Myocyte Enhancer Factor 2C). 
The BMD association has been independently replicated for four of the five genes: TNFRSF11B~\cite{Rivadeneira2009,Zhang2014,Kemp2014,Paternoster2013}, CTNNB1~\cite{Pei2016, Rivadeneira2009}, ZBTB40~\cite{Rivadeneira2009,  Zhang2014,Nielson2016}, and MEF2C~\cite{Pei2016a}. 
The association of IBSP has not been replicated in other GWAS studies, but homozygous knockouts of this gene have a known skeletal phenotype in mice. Mice deficient in IBSP have impaired bone growth and mineralization, with dramatically reduced bone formation~\cite{Malaval2008}. 
CTNBB1, MEF2C and TNFRSF11B also have known skeletal phenotypes. 
CTNNB1 is a key downstream member of the Wnt signaling pathway, and deletion of CTNNB1 in mice results in severe osteopenia, while constitutive activation of CTNNB1 results in massively increased bone deposition~\cite{Holmen2005}. 
MEF2C is a transcription factor which controls bone mass by regulating osteoclastic bone resorption, where mice with a MEF2C deletion in osteocytes had increased bone mass compared with normal controls~\cite{Kramer2012}. TNFRSF11B (or OPG) is a secreted glycoprotein which regulates bone mass~\cite{Simonet1997}. 
ZBTB40 has no known biological or molecular involvement in skeletal phenotypes~\cite{Styrkarsdottir2008}, even though it has been associated with BMD in multiple GWAS and is expressed in bone.
SNPs in linkage disequilibrium (LD) with each other at these loci also show a high importance score, as exemplified by TNFRSF11B in Figure 3b (Supplemental Figure 2 and 3 for the other genes). %~\ref{manhattenzoom}).

@@

In contrast, one of the lowest ranked gene with literature implied BMD association is DCDC5 (Doublecortin Domain Containing 5) ranked 23,842
(see Supplemental Figure 1). Similar to the conclusions drawn by Duncan {\it et
  al.}~\cite{Duncan.2011}, our results also indicate that this gene does not have a strong association with BMD. DCDC5 is not highly expressed in bone, and any mechanisms by which this gene might regulate BMD are unclear
\cite{Thakker2012}.  Although variants near DCDC5 are associated with BMD \cite{Rivadeneira2009}, it is questionable as
to whether the gene itself contributes to BMD.
Similarly, the lowest ranked gene with annotated BMD links, MAPT (microtubule associated protein tau) at 36,262, is primarily expressed in the nervous system and has not been identified in bone tissue. 
It is therefore likely that the association is through a co-morbidity with Alzheimer's disease~\cite{Dengler-Crish2017} and not a major contributing factor in this cohort.

\subsection{\cursedforest\ amplifies signal over traditional methods to identify novel associations}

Figure 1b shows a traditional GWAS analysis using logistic regression~\cite{Duncan.2011} where the resulting $p$-values do not clearly distinguish these known BMD genes from the noise. 
In contrast, \cursedforest\ seems to amplify the signal to identify associated features (Figure 1a). 
As such, this approach allows to employ a very stringent cut-off for identifying strongly associated feature. Here we chose a 99.98 False Discovery Rate (FDR) from 60 random experiments (see Methods).

@@

Above this FDR cut-off we identify four novel locations as annotated in Figure 1a.
%~\ref{figure:rbo-prod.png}. 
Most notably a lincRNA (LINC01006) on chromosome 7, which has been associated with serum creatinine levels in two independent GWAS studies~\cite{Pattaro2010,Chambers2010}. 
Subjects with low serum creatinine levels were at a higher risk for low BMD~\cite{Huh2015} and in a rat study high creatine monohydrate diet resulted in greater lumbar BMD~\cite{Antolic2007}, 
with creatinine being the breakdown product of creatine. The variants overlapping the 5' end of the lincRNA are the highest ranked features in this region and all LD SNPs also show high importance (see Fig 3d). Also amonst the top 30 features is PRODH (proline dehydrogenase) located on 22q11, a loci observed by Stagi {\it
  et al.}~\cite{Stagi.2010} to significantly reduce bone mass in individuals with a micro-deletion in this region. 
PRODH catalyses the first step in proline degradation, and as proposed by Phang {\it et al.}~\cite{Phang.2008}, proline is made available by collagen degradation. 
Again, all LD SNPs also show high importance (see Figure 3c). %~\ref{zoomplot}). 
%Ranked at position 5 (rs6961378) and 18 (rs146034056) are located on chromosome 7 overlapping the lincRNA (LINC01006), which as been linked to vertebrate development~\cite{Woolfe2005}.
In contrast rs17195090 and rs112558362 are single feature loci (Supplementary Figure 4) hence likely due to erroneous imputation or genotyping.

To evaluate whether the here identified features are significantly associated with BMD, we train a traditional linear mixed model to return a $p$-value for each included feature. Interestingly, while the three known BMD genes with the highest Importance score are indeed significantly associated (TNFRSF11B $p$-value=XX, CTNNB1 XX, ZBTB40 XX), IBSP and MEF2C were not significant. All of our newly associated features, on the other hand, were deemed significant, and in-fact rs17195090 was the most significant feature evaluated with $p$-value=XX (rs112558362 XX, LINC01006 XX, PRODH XX).

@@

%\cursedforest\ can successfully perform GWAS-type analyses replicating previous findings and uncovering novel associations due to random forests also modelling interacting features.

@@

%\begin{figure}[tbhp]
%  \begin{adjustwidth}{-2.00in}{0in}
%    \caption{\textbf{Rank of literature assigned BMD genes}}
%    \label{figure:knowngeneranks}
%    \includegraphics[totalheight=6.5cm]{./figs/BMDgenes_landscape.pdf}
%    \begin{flushleft}
%      The figure shows the rank of the 100 literature assigned BMD genes. Each ball represents a SNP assigned to the closest gene with colour representing the distance (max 100 kb). The size shows the Gini impurity score with larger balls representing more important SNPs. The horizontal line separates the genes with SNPs amongst the Top 20 associated features. 
%     \end{flushleft}
%  \end{adjustwidth}
%\end{figure}

\begin{figure}[tbhp]
  \begin{adjustwidth}{-2.00in}{0in}
    \caption{\textbf{Genome-wide importance of SNPs for Bone Mineral Density association}}
    \label{figure:ranksumtest}
     \begin{subfigure}[b]{0.5\linewidth}
     \includegraphics[totalheight=7cm]{./figs/BMDTopWholeGenome_annotated.png}
      \label{figure:BMDTopWholeGenome} 
    \end{subfigure} 
 \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[totalheight=7cm]{./figs/BMDTopWholeGenome_BW_random.png}
      \label{zoomplotTNFRSF} 
   \end{subfigure} 
    \begin{subfigure}[b]{0.5\linewidth}
    \includegraphics[totalheight=7cm]{./figs/new_10k_TNFRSF11B.pdf}
%     \includegraphics[totalheight=7cm]{./figs/new_10k_PRODH.pdf}
      \label{zoomplotprodh} 
    \end{subfigure} 
        \begin{subfigure}[b]{0.5\linewidth}
     \includegraphics[totalheight=7cm]{./figs/EUR_chr7_156_2-156_6mb.pdf}
      \label{zoomplotlinc} 
    \end{subfigure} 
    \begin{flushleft}
      Genome-wide SNPs and their importance for the BMD dataset {\bf(a)}; highlighted are the 100 literature assigned BMD genes (green) and four novel loci. The horizontal line indicates the top 99.98 FDR. 
      {\bf(b)} shows a previous GWAS analysis using logistic regression. 
      Locus Zoomplot for {\bf(c)} the top ranked gene TNFRSF11B and the two putative novel associations {\bf(d)} LINC01006. Each point represents a variant with the normalised Gini impurity score on the left y axis and LD to the central SNP in colour with the highest $r^2$ in red.
     \end{flushleft}
  \end{adjustwidth}
        \label{manhattenzoom} 
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Section 1:  Feature Selection and classification on synthetic data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest can evaluate variable importance and perform classification for millions of features}
%\label{section:synthetic}
In this section we test the limits of \cursedforest\'s association testing and classify ability on high-dimensional data. 
%In this section we demonstrate the two features of \cursedforest, namely: variable selection analysis to select the most
%associated set of features amongst millions of dimensions; and its ability to accurately classify the labels of
%high-dimensional samples.  
We therefore generate a synthetic dataset with 2.5 million variables ($p=2500000$), of which 5 
are related to the response variable, and 5000 samples ($n=5000$), as discussed in the methods section.

@@

Using a novel parallelization approach enabled by Spark, we extended random forests to cope with extremely high dimensional data (large number of variables). 
We have demonstrated that for datasets with more variable ($p$) than samples ($n$), such as GWAS applications, \cursedforest can perform effective variable selection replicating previous findings and uncovering novel associations in a Bone Mineral Density study. 
We are able to elevate the signal compared to previous approaches using logistic regression due to random forests also modelling interacting features.

In fact, \cursedforest's ability to almost exhaustively sample the combinatorial space by efficiently building large numbers of trees puts it closer than any other method to the theoretical boundary for accurate recovery of informative variables. Donoho and Tanner~\cite{Donoho.and.Tanner.2009} define this as the ``universal phase change'', which depends on underdetermination, $\delta = n/p$, and the sparsity, $\rho =k/n$ (where $k$ is the number of informative variables). 
By effectively operating in the difficult area of this space with $p >> n$ and small number of informative features, \cursedforest is pushing the practical limits of this paradigm.

To showcase the breadth of application scope, we also demonstrate that \cursedforest's implementation aids in classification problems such as predicting the ethnicity of the 1000 genome project. 
By comparing \cursedforest\ to other implementations (including those optimised for large datasets) we have demonstrated that runtime can be improved by 80\% allowing analysis in timeframes relevant for point-of-care diagnostics such as for "Patients like mine" scenarios thereby enabling evidence based self learning health care.
